---
title: "Multi-center Power Simulation"
author: Teeranan Pokaprakarn; Hiroto Udagawa; Danyi Xiong; Nusrat Rabbee (rabbee@berkeley.edu) 
date: Department of Statistics, University of California, Berkeley, CA
output: pdf_document
header-includes:
    - \usepackage{caption}
---
\captionsetup[table]{labelformat=empty}
```{r, echo=F}
library(knitr)
library(xtable)
install.packages('./SimHaz_0.1.tar.gz',repos = NULL, type="source")
library(survival)
```

##Multi-center Time-dependent Exposure

It is known that treatment effects can vary significantly across sites in multi-center trials. Variability due to a subject’s age or disease burden is different than that which arises due to center to center variability. For example, some centers and affiliated physicians are more aggressive in treating the disease under control while some are more conservative and prefer traditional treatment. In this method, we will study several methods for incorporating these center effects into our analysis of survival data. We will build on Glidden and Vittinghoff [1] and investigate their models and scenarios under time-dependent exposure. 

A multicenter design is often necessary to provide adequate sample and enhance generalizability of results. If these centers have clustering effects, ignoring them can create wrong and misleading results. Specifically using a simplistic pooled model (ignoring the centers or clusters) lead to underpowered studies and to potential biased estimates of treatment effects. In our presentation, we will introduce three different methods of accounting for these center effects: stratified Cox model, fixed effects Cox model, and the random effect frailty model.

* In the stratified Cox model, we take the hazard functions of each center to be unspecified. Thus this lack of structure makes this model the most general. This method essentially throws out all between-center comparisons and model each center based on within-center subjects. Thus as the number of centers increases with a fixed sample size, we lose more information. 

* The fixed effects Cox model assumes that the centers act proportionally on a single baseline hazard function. When the number of clusters is small relative to the sample size, this model tends to perform well.

* The frailty model treats the center effects as a sample from a probability distribution. The gamma frailty model assumes that this distribution is gamma with mean 1. The variance determines the between-cluster variability. 

We want to explore which method performs well when analyzing multi-center survival data with time-dependent exposure under various scenarios. According to Glidden and Vittinghoff [1], frailty models provides the most flexible, efficient framework for taking into account center effects in most settings.

We will run simulations to compare the performance of our two models with respect to power. We set our  = 0.7 for the simulations. We varied our settings based on total sample size (N) and center size.

##Example 1: Event rates across centers are equal

In Example 1 we run simulations where the event rates of all the centers (K=40) are equal. This is equivalent to having no center effect whatsoever. The user specifies the parameters by entering a data frame like the following. 


```{r, echo=T}
input_df <- data.frame(cat_id = as.factor(1:40), center.size = rep(10,40), 
                       cat_exp.prop = rep(1/3, 40), med.TTE.Control=rep(14,40)) 
```
```{r, echo=F}
kable(rbind(head(input_df,n=3),tail(input_df,n=3), make.row.names=F))
```

Once the parameters for the centers are specified, we can pass this into getpower.multicenter(). To test the different methods, we can specify the ‘method’ argument of the function.  For example, a simulation using a stratified cox model would be specified using the following function call:

```{r, eval=FALSE}
getpower.multicenter(nSim = 500, N = 100, beta = 0.7, df = input_df, 
                     method="strata", dist=NULL, type = "td", 
                     scenario = "strata", maxrelexptime = 1/6, min.futime =4,         
                     min.postexp.futime = 4, output.fn = "output_mult.csv")
```


We ran this simulation with three different methods: Model with Independence Assumption, Strata, and Frailty. 

We then varied the number of total subjects and the center size which also affects the number of centers in our models. The results of our simulations are shown in the table below. 

```{r, cache=T, echo=F}
#scenario 1, no center effect

input_df1 <- data.frame(cat_id = as.factor(1:10), center.size = rep(10,10), 
                        cat_exp.prop = rep(1/3, 10),
                        med.TTE.Control=rep(14,10))
table1 = data.frame()

for(i in c('Model with Independence Assumption','strata','frailty')){
    table1 <- rbind(table1,getpower.multicenter(nSim = 500, N = 100, beta = 0.7, df = input_df1, method=i,dist=NULL, type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4,min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}


input_df2 <- data.frame(cat_id = as.factor(1:5), center.size = rep(20,5), 
                        cat_exp.prop = rep(1/3, 5),
                        med.TTE.Control=rep(14,5))
table2 = data.frame()

for(i in c('Model with Independence Assumption','strata','frailty')){
    table2 <- rbind(table2,getpower.multicenter(nSim = 500, N = 100, beta = 0.7, df = input_df2,method=i, type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}


input_df3 <- data.frame(cat_id = as.factor(1:40), center.size = rep(10,40), 
                        cat_exp.prop = rep(1/3, 40),
                        med.TTE.Control=rep(14, 40))
table3 = data.frame()

for(i in c('Model with Independence Assumption','strata','frailty')){
    table3 <- rbind(table3,getpower.multicenter(nSim = 500, N = 400, beta = 0.7, df = input_df3,method=i, type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}

input_df4 <- data.frame(cat_id = as.factor(1:20), center.size = rep(20,20), 
                        cat_exp.prop = rep(1/3, 20),
                        med.TTE.Control=rep(14, 20))
table4 = data.frame()

for(i in c('Model with Independence Assumption','strata','frailty')){
    table4 <- rbind(table4,getpower.multicenter(nSim = 500, N = 400, beta = 0.7, df = input_df4,method=i, type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}
```

```{r xtable, results='asis',echo=F}
final1 = data.frame(rbind(table1,table2,table3,table4))
colnames(final1) = c('Method','Center Size','N','Beta','Power')
print(xtable(final1, caption = "Table 1: Equal Event Rates",,digits=c(0,0,0,0,5,3)), comment=F, hline.after = c(-1,0,3,6,9,12), booktabs=T,include.rownames = FALSE)
```

##Example 2: Events rates across centers are different

Example 2a runs simulations where the median time to event (control group) for each center is specified by the user. For our simple case below (K=3), we create a table like below:

```{r}
input_df <- data.frame(cat_id = as.factor(1:3), center.size = rep(100,3), 
                       cat_exp.prop = rep(1/3, 3), med.TTE.Control=c(14,20,31)) 
```
```{r echo = F}
kable(input_df)
```

We run the simulation in the same way we ran scenario 1 but with our new input data frame. 

We do this now with four different methods: marginal cox, fixed effects, stratified, and frailty. Unlike scenario 1, we keep the number of centers fixed at three (K=3) and run four methods:

```{r, cache=T, echo=F}
#Scenario 2
input_df5 <- data.frame(cat_id = c("low","med","high"), center.size = rep(100,3), 
                        cat_exp.prop = rep(1/3, 3),
                        med.TTE.Control=c(14,20,31))

table5 = data.frame()

for(i in c('marginal','fixed effects','strata','frailty')){
    table5 = rbind(table5, getpower.multicenter(nSim = 500, N = 300, beta = 0.7, df = input_df5,method=i, type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}
```

```{r, results='asis', echo=F}
final2 = data.frame(table5)
colnames(final2) = c('Method','Center Size','N','Beta','Power')

print(xtable(final2, caption = "Table 2: Fixed Different Event Rates for 3 Centers",digits=c(0,0,0,0,5,3)), comment=F,booktabs=T,include.rownames = FALSE)

```

For this scenario, the marginal model will act as our reference standard. Based on our results, we see while all of the models overestimates beta, frailty performs better than our fixed effect and stratified models. 

##Example 3: Center effects are gamma distributed


In our final example, we mimic the method described in the Glidden and Vittinghoff [1] paper where we generate data for center effects following the Gamma distribution. We investigate the method using two examples.

In the first we set the same baseline median time to event (control group) for each center. The data frame below shows an N value of 200, a center size of 10 (M=10), number of clusters is 20 (K=20).

```{r}
input_df <- data.frame(cat_id = as.factor(1:20), center.size = rep(10,20), 
                        cat_exp.prop = rep(1/3, 20),
                        med.TTE.Control=rep(14,20)) 
```
```{r, echo=F}
kable(rbind(head(input_df,n=3),tail(input_df,n=3),make.row.names=F))
```

Once the data frame is created, we can pass it into getpower.multicenter(). Like in the previous scenarios, we can specify the ‘method’ argument of the function for the ‘marginal’, ‘strata’, and ‘frailty’ models.  Note, however, that we now add the ‘dist = “gamma”’ argument in order to add the random center effect sampled from a gamma distribution.

```{r, eval=F}

getpower.multicenter(nSim = 500, N = 100, beta = 0.7, df = input_df, 
                     method="strata", dist= "gamma", type = "td", 
                     scenario = "strata", maxrelexptime = 1/6, min.futime =4,         
                     min.postexp.futime = 4, output.fn = "output_mult.csv")

```

We run this simulation for our ‘marginal’, ‘strata’, and ‘frailty’ methods with varying center sizes, and N values.

```{r, cache=T, echo=F}
input_df7 <- data.frame(cat_id = as.factor(1:20), center.size = rep(10,20), 
                        cat_exp.prop = rep(1/3, 20),
                        med.TTE.Control=rep(14,20))

table7 = data.frame()

for(i in c('strata','frailty')){
    table7 <- rbind(table7, getpower.multicenter(nSim = 500, N = 200, beta = 0.7, df = input_df7,method=i, dist="gamma", type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}


input_df8 <- data.frame(cat_id = as.factor(1:10), center.size = rep(20,10), 
                        cat_exp.prop = rep(1/3, 10),
                        med.TTE.Control=rep(14,10))

table8 = data.frame()

for(i in c('strata','frailty')){
    table8 <- rbind(table8, getpower.multicenter(nSim = 500, N = 200, beta = 0.7, df = input_df8,method=i, dist="gamma", type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}


input_df9 <- data.frame(cat_id = as.factor(1:40), center.size = rep(10,40), 
                        cat_exp.prop = rep(1/3, 40),
                        med.TTE.Control=rep(14,40))
table9 = data.frame()

for(i in c('strata','frailty')){
    table9 <- rbind(table9,getpower.multicenter(nSim = 500, N = 400, beta = 0.7, df = input_df9,method=i, dist="gamma",type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4, min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}


input_df10 <- data.frame(cat_id = as.factor(1:20), center.size = rep(20,20), 
                         cat_exp.prop = rep(1/3, 20),
                         med.TTE.Control=rep(14,20))
table10 = data.frame()

for(i in c('strata','frailty')){
    table10 = rbind(table10, getpower.multicenter(nSim = 500, N = 400, beta = 0.7, df = input_df10,method=i , dist="gamma",type = "td", scenario = i, maxrelexptime = 1/6, min.futime = 4,min.postexp.futime = 4)[1,c(1,8,3,17,24)])
}
```

```{r, results='asis',echo=F}
final3 = data.frame(rbind(table7,table8,table9,table10))
colnames(final3) = c('Method','Center Size','N','Beta','Power')

print(xtable(final3, caption = "Table 3: Gamma Centered Event Rates",digits=c(0,0,0,0,5,3)), comment=F, hline.after = c(-1,0,2,4,6,8), booktabs=T,include.rownames = FALSE)

```


In this example, we generated the center effects from a gamma distribution with shape parameter of 2 and scale parameters of 0.5. This distribution has a mean of 1, which will form centers with frailty > 1 and frailty < 1. This simulates our high risk and low risk groups respectively. 

In this scenario, the frailty model is recommended over the stratified model since we used gamma frailty to generate the data and ceter effects. 

##References

[1] Glidden D, Vittinghoff E. Modelling clustered survival data from multicentre clinical trials. STATISTICS IN MEDICINE 2004;
